---
title: "Database Connections in R"
author: "Susan VanderPlas"
date: "March 7, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Notes on how to connect to various databases using R. 

## Connecting to Oracle databases
* Ensure Oracle is installed on the machine. Locate the oracle install directory, hereafter referred to as "~\\Oracle\\"

* Set the following environment variables (Right click on My Computer -> Properties -> Advanced System Settings -> Environment Variables):
    * OCI_INC: "~\\Oracle\\oci\\include"
    * (PICK ONE) OCI_LIB32: "~\\Oracle\\bin" (if you have 32-bit oracle installed)
    * (PICK ONE) OCI_LIB64: "~\\Oracle\\bin" (if you have 64-bit oracle installed)
    
* Install ROracle: `install.packages("ROracle", type = "source")`

* To connect to a specific database, you will need to look in the TNSNAMES.ORA file, which should be somewhere in the Oracle installation directory. The following code is a skeleton for how to connect to an oracle database. 

```{r r-oracle-conn, eval = F}

library(ROracle)

drv <- dbDriver("Oracle")

connect.string <- "(DESCRIPTION =
    (ADDRESS_LIST =
       (ADDRESS = (COMMUNITY = ...)(PROTOCOL = TCP)(Host = ...)(Port = ...))
    )
  (CONNECT_DATA =
      (INSTANCE_NAME = PROD)
    (SERVICE_NAME = PROD.WORLD)
  )
)" # From the TNSNAMES.ORA file

# Actually connect
con <- dbConnect(drv, "your username", "your password",
                 dbname = connect.string)

# Queries:
tmp <- dbGetQuery(con, "Query text here")
```

## Connecting to SQL databases
* Install RODBC: `install.packages("RODBC")`

* Here is skeleton code for how to connect to a SQL database:

```{r r-sql-conn, eval = F}
con <- odbcDriverConnect(
  'driver={SQL Server};
  server=server_name;
  database=db_name;
  trusted_connection=true' # This uses windows login information 
                           # instead of a username/password
) 

# Queries:
tmp <- sqlQuery(con, "Query text here")
```
