---
title: "Graphics in R"
output: 
    revealjs::revealjs_presentation:
        transition: fade
        theme: white
        highlight: zenburn
        css: ../css/slides.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, comments = T, collapse = T)
```

```{r, echo=FALSE}
library(ggplot2)
```

# Basic Plots

## Setup

Run the [Setup.R](./code/Setup.R) file.

If everything works correctly, you should see a plot:
```{r, echo = F, fig.width = 7, fig.height = 4}
qplot(rnorm(100), rnorm(100), colour = rbinom(100, 5, 0.2))
```

## ggplot2 In a Nutshell

- Package for statistical graphics
- Developed by Hadley Wickham
- Designed to adhere to good graphical practices
- Supports a wide variety plot types
- Constructs plots using the concept of layers\medskip
- http://had.co.nz/ggplot2/ or Hadley's book *ggplot2: Elegant Graphics for Data Analysis}* for reference material


## *qplot* Function

The *qplot()* function is the basic workhorse of ggplot2

- Produces all plot types available with ggplot2
- Allows for plotting options within the function statement
- Creates an object that can be saved
- Plot layers can be added to modify plot complexity

## *qplot* Structure

The *qplot()* function has a basic syntax:

*qplot(variables, plot type, dataset, options)*

- variables: list of variables used for the plot
- plot type: specified with a *geom =* statement
- dataset: specified with a *data =* statement
- options: there are so, so many options!


## Diamonds Data

Objective: Explore the diamonds data set (preloaded along with ggplot2) using `qplot` for basic plotting.

The data set was scraped from a diamond exchange company data base.  It contains the prices and attributes of over 50,000 diamonds.


## Examining the Diamonds Data

What does the data look like?

Look at the top few rows of the diamond data frame to find out!

```{r}
head(diamonds)
```


## Basic Scatterplot

Basic scatter plot of diamond price vs. carat weight

```{r, fig.height=4, fig.width=7}
qplot(carat, price, geom = "point", data = diamonds)
```

## Another Scatterplot

Scatter plot of diamond price vs carat weight showing versitility of options in `qplot`

```{r, fig.height=3, fig.width=7}
qplot(carat, log(price), geom = "point", data = diamonds, 
	alpha = I(0.2), colour = color, 
	main = "Log price by carat weight, grouped by color") + 
    xlab("Carat Weight") + ylab("Log Price")
```

## Your Turn {data-background=#527a7a}

All of the your turns for this section will use the tips data set:

```{r}
tips <- read.csv("http://srvanderplas.github.io/NPPD-Analytics-Workshop/02.Graphics/data/tips.csv")
```

1. Use qplot to build a scatterplot of variables tips and total bill
2. Use options within qplot to color points by smokers
3. Clean up axis labels and add main plot title
