---
title: "Data Manipulation in R"
output: 
  revealjs::revealjs_presentation:
    includes:
      in_header: header.html
    transition: fade
    theme: white
    highlight: zenburn
    reveal_options:
      slideNumber: true
      previewLinks: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, comments = T, collapse = T, cache = T)
```

# Reading in Data

## Setup

```{r, eval = T}
# Setup.R file
source("https://bit.ly/2i8sicn")
```

## Data in Excel

Formats xlsx and csv - what's the difference?

- File extensions xls and xlsx are proprietary Excel formats, binary files
- csv is an extension for Comma Separated Value files. They are text files - directly readable.

- Example: Gas prices in midwest since 1994


## Gas Prices in the Midwest

```{r, echo = 2:3}
midwest <- read.csv("https://srvanderplas.github.io/NPPD-Analytics-Workshop/03.DataManipulation/data/midwest.csv")
midwest <- read.csv("https://bit.ly/2hwWiQ8")
head(midwest)
```


## Gas Prices

```{r}
str(midwest)
```

There is clearly some work to be done...

***
### `read.csv` vs. `read.table`

`read.csv` is really just a wrapper for `read.table` with certain parameters set:

```{r}
read.csv
```

So to properly read in this data, it's probably best to use `read.table` directly

```{r, eval=FALSE}
?read.table
```


## Your Turn {.yourturn  data-background=#527a7a}

Have a look at the parameters of read.table (?read.table) to solve the following problems:

1. Read the first two lines of the file into an object called `midwest_names`
2. Read everything EXCEPT the first two lines into an object called `midwest_data`

```{r, echo=FALSE}
midwest_names <- read.table("https://bit.ly/2hwWiQ8", nrows = 2, sep = ",", stringsAsFactors = FALSE)
midwest_data <- read.table("https://bit.ly/2hwWiQ8", skip = 2, sep = ",", stringsAsFactors = FALSE)
```

## Solutions {.solutions  data-background=#527a7a}
1. Read the first two lines of the file into an object called `midwest_names`
```{r}
midwest_names <- read.table("https://bit.ly/2hwWiQ8", 
                            nrows = 2, sep = ",", 
                            stringsAsFactors = FALSE)
```

2. Read everything EXCEPT the first two lines into an object called `midwest_data`
```{r}
midwest_data <- read.table("https://bit.ly/2hwWiQ8", 
                           skip = 2, sep = ",", 
                           stringsAsFactors = FALSE)
```


## Demo: Data Cleaning

```{r, warning=FALSE, message=FALSE}
values <- c(midwest_data$V3, midwest_data$V5, midwest_data$V7, 
            midwest_data$V9, midwest_data$V11)
dates <- c(paste(midwest_data$V1, midwest_data$V2, sep = "-"), 
           paste(midwest_data$V1, midwest_data$V4, sep = "-"),
           paste(midwest_data$V1, midwest_data$V6, sep = "-"),
           paste(midwest_data$V1, midwest_data$V8, sep = "-"),
           paste(midwest_data$V1, midwest_data$V10, sep = "-"))

dates <- dates[!is.na(values)]
values <- values[!is.na(values)]

library(lubridate)
dates <- ymd(dates)

midwest_gas <- data.frame(date = dates, price = values)
midwest_gas <- midwest_gas[with(midwest_gas, order(date)), ]
```

\*\* A better way to do this will be shown later \*\*

## Make a plot
```{r, warning=FALSE, message=FALSE, fig.width=7, fig.height=4}
library(ggplot2)
qplot(date, price, data = midwest_gas, geom = "line")
```


## Reading Excel Data

```{r, eval=FALSE}
library(readxl)
download.file(url = "http://bit.ly/2ihV8Ye", 
              destfile = "midwest.xls")

midwest2 <- read_excel("midwest.xls")
# Fix name issues
names(midwest2) <- make.names(names(midwest2))
head(midwest2, 4)
```

```{r, echo=FALSE}
library(readxl)

midwest2 <- read_excel("./data/midwest.xls")
names(midwest2) <- make.names(names(midwest2))
head(midwest2, 4)
```


## `foreign` Package

Other file formats can be read using the functions from package `foreign`

- SPSS: `read.spss`
- SAS: `read.xport`
- Minitab: `read.mtp`


## Your Turn {.yourturn  data-background=#527a7a}

The NHANES (National Health and Nutrition Survey) publishes data in the SAS xport format:

http://wwwn.cdc.gov/Nchs/Nhanes/Search/nhanes13_14.aspx

1. Scroll to the bottom, choose one of the datasets (Demographics, Examination, etc.). Download the Data file (XPT)
2. Use `read.xport()` to load the file into R

## Solutions {.solutions  data-background=#527a7a}
```{r, eval = F}
download.file("https://wwwn.cdc.gov/Nchs/Nhanes/2013-2014/DEMO_H.XPT", "DEMO.XPT")
library(foreign)
demographic.vars <- read.xport("DEMO.XPT")
```

## Your Turn {.yourturn  data-background=#527a7a}

MINITAB publishes several sample data sets. Download the Paint hardness data

http://support.minitab.com/en-us/datasets/PaintHardness.MTW

and read it in to R.

## Solutions {.solutions  data-background=#527a7a}
```{r}
# download.file("http://support.minitab.com/en-us/datasets/PaintHardness.MTW", "paint.MTW")
# library(foreign)
# demographic.vars <- read.mtp("paint.MTW")
```

# Summarizing Data

## Baseball Data

- The `plyr` package contains the data set `baseball`
- seasonal batting statistics of all major league players (through 2007)
 
```{r}
data(baseball, package = "plyr")
help(baseball, package = "plyr")
head(baseball)
```

## Baseball Data

- Goal: create career summary statistics for each player
- Plan: subset on a player, and compute statistics

```{r}
ss <- subset(baseball, id == "sosasa01")
head(ss, 3)
mean(ss$h / ss$ab)
```

There should be an **automatic** way to calculate this.


## `for` loops

- Idea: repeat the same (set of) statement(s) for each element of an index set
- Setup: 
    - Introduce counter variable (sometimes named `i`)
    - Reserve space for results
- Generic Code:

```{r, eval=FALSE}
result <- rep(NA, length(indexset))
for (i in indexset) {
  ... some statements ...
  result[i] <- ...
}
```
    
## `for` loops for Baseball

- Index set: player id
- Setup: 

```{r}
players <- unique(baseball$id)
n <- length(players)

ba <- rep(NA, n)

for (i in 1:n) {
  career <- subset(baseball, id == players[i])
  ba[i] <- with(career, mean(h / ab, na.rm = TRUE))
}

summary(ba)
```

## `for` loops for Baseball

- Index set: player id
- `i = 0`: 

```{r}
players <- unique(baseball$id)
n <- length(players)

ba <- rep(NA, n)

head(ba)
```

## `for` loops for Baseball

- Index set: player id
- `i = 1`: 

```{r}
players <- unique(baseball$id)

ba <- rep(NA, length(players))

for (i in 1:1) {
  career <- subset(baseball, id == players[i])
  ba[i] <- with(career, mean(h / ab, na.rm = TRUE))
}

head(ba)
```

## `for` loops for Baseball

- Index set: player id
- `i = 2`: 

```{r}
players <- unique(baseball$id)

ba <- rep(NA, length(players))

for (i in 1:2) {
  career <- subset(baseball, id == players[i])
  ba[i] <- with(career, mean(h / ab, na.rm = TRUE))
}

head(ba)
```


## `for` loops for Baseball

- Index set: player id
- `i = 3`: 

```{r}
players <- unique(baseball$id)

ba <- rep(NA, length(players))

for (i in 1:3) {
  career <- subset(baseball, id == players[i])
  ba[i] <- with(career, mean(h / ab, na.rm = TRUE))
}

head(ba)
```


## Your Turn {.yourturn data-background=#527a7a}

- MLB rules for the greatest all-time hitters are that players have to have played 
    - at least 1000 games 
    - with at least 1000 at-bats in order to be considered
- Extend the for loop above to collect the additional information
- Introduce and collect data for two new variables: `games` and `atbats`

## Solution {.solutions data-background=#527a7a}

```{r}
library(dplyr)
data(baseball, package = "plyr")


players <- unique(baseball$id) 
ba <- rep(NA, length(players)) 
games <- rep(NA, length(players)) 
atbats <- rep(NA, length(players)) 
for (i in 1:length(players)) { 
  career <- subset(baseball, id == players[i]) 
  ba[i] <- with(career, mean(h / ab, na.rm = TRUE)) 
  games[i] <- with(career, sum(g, na.rm = TRUE))
  atbats[i] <- with(career, sum(ab, na.rm = TRUE))
}
	
```


***
### How did it go? What was difficult?

- Household chores (declaring variables, setting values each time) distract from real work
- Indices are error-prone
- Loops often result in slow code because R can compute quantities using entire vectors in an optimized way

## Summarise

A special function: `summarise` or `summarize`

```{r, message=FALSE}
library(dplyr)

baseball <- read.csv("https://bit.ly/2iFyIwL")

summarise(baseball,
          ba = mean(h / ab, na.rm = TRUE),
          games = sum(g, na.rm = TRUE),
          hr = sum(hr, na.rm = TRUE),
          ab = sum(ab, na.rm = TRUE))
```

## More Summarise

```{r}
summarise(subset(baseball, id == "sosasa01"), 
          ba = mean(h / ab, na.rm = TRUE),
          games = sum(g, na.rm = TRUE),
          hr = sum(hr, na.rm = TRUE),
          ab = sum(ab, na.rm = TRUE))
```

## `dplyr` + `Summarise`

- A powerful combination to create summary statistics

```{r}
careers <- summarise(group_by(baseball, id),
                 ba = mean(h / ab, na.rm = TRUE),
                 games = sum(g, na.rm = TRUE),
                 homeruns = sum(hr, na.rm = TRUE),
                 atbats = sum(ab, na.rm = TRUE))

head(careers)
```

## Pipes
Rather than nesting functions(inside(each(other))), use a pipe `%>%`:
```{r}
careers <- baseball %>%
  group_by(id) %>%
  summarize(
    ba = mean(h / ab, na.rm = TRUE),
    games = sum(g, na.rm = TRUE),
    homeruns = sum(hr, na.rm = TRUE),
    atbats = sum(ab, na.rm = TRUE)
  )
```

Pipes make code more readable and easier to understand.

## Pipes
`a %>% function(b)`    
is the same as     
`function(a, b)`:

the piped variable is substituted for the first function argument. 

## Your Turn {.yourturn data-background=#527a7a}

- Find some summary statistics for each of the teams (variable `team`)
    - How many different (`unique`) players has the team had?
    - What was the team's first/last season?
- Challenge: Find the number of players on each team over time. Does the number change?


## Solutions {.solutions data-background=#527a7a}
1. Find some summary stats for each of the teams (`team`)
    - How many `unique` players has the team had?
    - What was the team's first/last season?
```{r}
baseball %>% 
  group_by(team) %>% 
  summarise(
    nplayer = length(unique(id)), 
    first = min(year), 
    last = max(year))
```


## Solutions {.solutions data-background=#527a7a}
2. Challenge: Find the number of players on each team over time. Does the number change?

```{r, echo = 1, fig.width = 7, fig.height = 3}
players.per.year <- baseball %>% 
  group_by(team, year) %>% 
  summarize(nplayer = n()) 

ggplot(data = players.per.year) + 
  geom_point(aes(x = year, y = nplayer, color = team)) + 
  geom_smooth(aes(x = year, y = nplayer)) + 
  scale_color_discrete(guide = F)
```
