---
title: "Introduction to R"
output:
  html_document:
    code_folding: show
    fig_caption: yes
    self_contained: yes
    theme: cerulean
    toc: yes
  word_document:
    reference_docx: U:\Templates\Rmarkdown Word\word_template.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = T, error = T,
                      tidy.opts = list(width.cutoff = 50))
```

There are many software programs designed to store, summarize, visualize, or analyze data:

- Spreadsheets: Microsoft Excel, Google Sheets, etc.
- Data Analysis and Statistical programs: SAS, SPSS, Minitab
- Programming languages: Python, C, C++, Fortran, Java
- Database software: Microsoft Access, SAP, SQL databases

`R` is statistical software which can read data, manipulate data, summarize data, fit statistical models, create quality plots, and more. It is used by companies like Google, Twitter, and Microsoft, as well as in government agencies and at national labs. 

# Setting Up

This guide will help set up R and RStudio on your machine. It is included for reference. 

## Installing R:

On Windows or OS X:

- Go to http://www.r-project.org/
- Click the CRAN link on the left, and pick a download site (0-Cloud is a good choice)
- Choose link based on your OS
- On Windows, choose the "base" subdirectory to install R.
- On OS X, choose the .pkg file to install R.

## Installing RStudio:

- Browse to https://www.rstudio.com/
- Mouse over Products and click RStudio
- Choose RStudio Desktop
- Click Download RStudio Desktop
- Choose the installer appropriate for your platform

# Example: Tips data

## Goals

- Explore a real dataset using R
- Get the "flavor" of R for data management and exploration
- Don't focus on the code - it will be explained later and in much more detail

## Tips Dataset

A server recorded the tips they received over about 10 weeks, including several variables:

  - Amount they were tipped
  - Cost of the total bill
  - Characteristics about the party (# people, gender, etc.)

Primary Question: How do these variable influence the amount being tipped?

Follow along using [Tips-Example.R](code/Tips-Example.R)

## First look: Reading in the data

Lets use R to look at the top few rows of the tips data set. First, we load tip using `read.csv`:

```{r}
tips <- read.csv("https://srvanderplas.github.io/NPPD-Analytics-Workshop/01.Introduction/data/tips.csv")
```

Now, we use the `head` function to look at the first 6 rows of the data:

```{r}
head(tips)
```

How big is this data set and what types of variables are in each column?

```{r}
str(tips)
```

We can get a summary of the values for each variable in tips:

```{r}
summary(tips)
```


What is the relationship between total bill and tip value?
First, we need to install and load `ggplot2` (a graphics library):

```{r, eval=FALSE}
install.packages("ggplot2")
library(ggplot2)
```

```{r, echo=FALSE, message = F, warning = F}
library(ggplot2)
```

```{r, fig.height = 3, fig.width = 6}
qplot(x = tip, y = total_bill, geom = "point", data = tips)
```

We can add linear regression line to the plot:

```{r, fig.height = 4, fig.width = 6}
qplot(x = tip, y = total_bill, geom = "point", data = tips) + 
    geom_smooth(method = "lm")
```

Tipping is generally done using a rule of thumb based on a percentage of the total bill. We will make a new variable in the data set for the tipping rate = tip / total bill

```{r}
tips$rate <- tips$tip / tips$total_bill

summary(tips$rate)
```

What is the distribution of tip rate in this dataset? We can use R to create a histogram:

```{r, fig.height = 4, fig.width = 6}
qplot(x = rate, data = tips, binwidth = .01)
```

One person tipped over 70%, who are they?

```{r}
tips[which.max(tips$rate),]
```

We can also calculate the average tipping rate for men and women separately:

```{r}
mean(tips$rate[tips$sex == "Male"])
mean(tips$rate[tips$sex == "Female"])
```

Is the difference between the two groups statistically significant? We can calculate a t-test to examine whether the average tip rate for men is the same as the average tip rate for women. 

```{r}
t.test(rate ~ sex, data = tips)
```

The p-value is greater than 0.05, and the confidence interval contains 0: there is no evidence of a significant difference in tipping rate between men and women. 

We can compare two groups using a side-by-side boxplot. For instance, we can see whether smokers tip at a different rate than non-smokers: 

```{r, fig.height = 3, fig.width = 6}
qplot(x = smoker, y = rate, geom = "boxplot", data = tips)
```

## Your Turn!

Try playing with chunks of code shown above to further investigate the tips data:

1. Get a summary of the total bill values
2. Make side by side boxplots of tip rates for different days of the week
3. Find the average tip value for smokers

### Solutions

1. Get a summary of the total bill values
```{r}
summary(tips$total_bill)
```

2. Make side by side boxplots of tip rates for different days of the week
```{r}
qplot(day, rate, geom = "boxplot", data = tips)
```

3. Find the average tip value for smokers
```{r}
mean(tips$tip[tips$smoker == "Yes"])
```

# R Basics

## Getting Help in R
The `help()` function is useful for getting help with a function:

```{r, eval = F}
help(head)
```

The `?` function also works:

```{r, eval = F}
?head
```

When searching for results online, it is helpful to use R + CRAN + \<query\> to get good results.

## R Reference Card

A copy of the R reference card is available at: 

http://cran.r-project.org/doc/contrib/Short-refcard.pdf

This card contains short versions of the most common functions used in R.

## R as an Overgrown Calculator

R can perform simple mathematical operations.
```{r}
# Addition and Subtraction
2 + 5 - 1

# Multiplication
109*23452

# Division
3/7
```


Here are a few more complex operations:
```{r}
# Integer division
7 %/% 2

# Modulo operator (Remainder)
7 %% 2

# Powers
1.5 ^ 3
```

```{r}
# Exponentiation
exp(3)

# Logarithms
log(3)
log(3, base = 10)

# Trig functions
sin(0)
cos(0)
tan(pi/4)

asin(0)
acos(1)
atan(1)
```

## Variables in R

Variables in R are created using the assignment operator, `<-`:

```{r}
x <- 5
R_awesomeness <- Inf
MyAge <- 21 #Haha
```

These variables can then be used in computation:
```{r}
log(x)
MyAge ^ 2
```

### Rules for Variable Names

- Variable names can't start with a number
- Names are case-sensitive
- Common letters are used internally by R and should be avoided as variable names    
`c, q, t, C, D, F, T, I`
- There are reserved words that R won't let you use for variable names.    
`for, in, while, if, else, repeat, break, next`
- R will let you use the name of a predefined function.    
Try not to overwrite those!

#### Error messages 
```{r}
# Variable starts with a number
1age <- 3
```

```{r}
# Case Sensitive
Age <- 3
age
```

```{r}
# Special Words can't be variable names
for <- 3
```

#### A Cautionary Tale
```{r, eval = T}
# This is a VERY bad idea:

T <- FALSE
F <- TRUE

T == FALSE
F == TRUE

rm(T, F) # Fix it!

T == FALSE
```
Note: In R, `T` and `F` are shorthand for `TRUE` and `FALSE` respectively. 

## Vectors
A variable can contain more than one value.    
A _vector_ is a variable which contains a set of values of the same type.    
The `c()` (combine) function is used to create vectors:

```{r}
y <- c(1, 5, 3, 2)
z <- c(y, y)
```

R performs operations on the entire vector at once:
```{r}
y / 2
z + 3
```

Vectors can be created using the `c()` or `rep()` function. To create a vector of consecutive values, use the `:` function:

```{r}
a <- 10:15
a
```

### Modifying Vectors

Vectors can be modified using indexing:

```{r}
# Get the total bill out of the tips dataset
bill <- tips$total_bill

x <- bill[1:5]
x
x[1] <- 20
x
```

Elements of a vector must all be the same type:

```{r}
head(bill)
bill[5] <- ":-("
head(bill)
```

By changing a value to a string, all the other values were changed to strings as well.

### Your Turn

Using the R Reference Card (and the Help pages, if needed), do the following:

1. Find out how many rows and columns the `iris' data set has.  Figure out at least 2 ways to do this.    
*Hint: "Variable Information" section on the first page of the reference card!*
2. Use the `rep` function to construct the following vector:  `1 1 2 2 3 3 4 4 5 5`    
*Hint: "Data Creation" section of the reference card*
3. Use `rep` to construct this vector: `1 2 3 4 5 1 2 3 4 5 1 2 3 4 5`

#### Solutions

1. Find out how many rows and columns the `iris' data set has.  Figure out at least 2 ways to do this.

```{r, collapse = T}
data(iris)

# first way: 
nrow(iris)
ncol(iris)

# second way: 
dim(iris)

# third way: 
str(iris) # look at the top line
```

2. Use the `rep` function to construct the following vector:  1 1 2 2 3 3 4 4 5 5
```{r}
rep(c(1:5), each = 2)
```

3. Use `rep` to construct this vector: 1 2 3 4 5 1 2 3 4 5 1 2 3 4 5
```{r}
rep(c(1:5), times = 3)
```

Vectors can be manipulated by indexing. 

Different elements of a vector can be extracted using brackets:

```{r}
a[1]
a[2]
a[5]
```

Indexes can also be more complicated: 
```{r}
a[c(1, 3, 5)]
a[1:5]
```

## Logical Values

- R has built in support for logical values
- TRUE and FALSE are built in.  T (for TRUE) and F (for FALSE) are supported but can be modified
- Logicals can result from a comparison using
    - $<$
    - $>$
    - $<=$
    - $>=$
    - $==$
    - $!=$
    
### Indexing with Logicals

Logical vectors can be used for indexing as well:

```{r}
x <- c(2, 3, 5, 7)
x[c(TRUE, FALSE, FALSE, TRUE)]
x > 3.5
x[x > 3.5]
```

#### Examples

```{r}
# Get the rate variable out of the tips dataset
rate <- tips$rate 

head(rate)

sad_tip <- rate < 0.10

rate[sad_tip]
```

## Data Frames

A collection of vectors, similar to a table in an Excel spreadsheet

- A data set is stored in a data frame
- Each column is a vector of the same length
- Each column can be a different type of data
- Each element in the vector/column has to have the same type of data
- columns can be accessed using `$`


`tips` is a data frame:
```{r}
head(tips)
```

`tips$sex` shows the `sex` column of `tips`
```{r}
tips$sex[1:20]
# Show the first 20 items in the sex column of tips
```


### Your Turn

1. Find out how many people tipped over 20%.    
*Hint: use the `sum` function on a logical vector to calculate how many TRUEs are in the vector:*
```{r}
sum(c(TRUE, TRUE, FALSE, TRUE, FALSE))
```
2. **More Challenging**: Calculate the sum of the total bills of anyone who tipped over 20%

#### Solutions
1. How many people tipped over 20%?

```{r}
sum(tips$rate > .2)
```

2. Sum of the total bills where the tip was over 20%

```{r}
sum(tips$total_bill[tips$rate > .2])
```


## Data Types in R

- Can use `mode` or `class` to find out information about variables
- `str` is useful to find information about the structure of your data
- Many data types: numeric, integer, character, Date, and factor most common
    
```{r}
str(tips)
```

Use `mode` and `class` to get more general information:
```{r}
class(tips)
mode(tips)
```
tips is a data frame, which is a list of vectors that have the same length. 

### Converting Between Types

Convert variables to a different type using the `as` series of functions:

```{r}
size <- head(tips$size)
size
as.character(size)
as.numeric("2")
```


## Some useful functions

There are a whole variety of useful functions to operate on vectors. 

```{r}
tip <- tips$tip
x <- tip[1:5]
length(x) # Number of elements of a vector
sum(x) # Sum of elements in a vector
```

Using the basic functions it wouldn't be hard to compute some basic statistics.

```{r}
# Homemade Statistical Functions
(n <- length(tip))
(meantip <- sum(tip) / n)
(standdev <- sqrt(sum((tip - meantip)^2) / (n - 1)))
```

But these functions are already built in to R.

```{r}
# Built in statistical functions
mean(tip)
sd(tip)
summary(tip)
quantile(tip, c(.025, .975))
```

## Element-wise Logical Operators

- `&` (elementwise AND)
- `|` (elementwise OR)

```{r}
c(T, T, F, F) & c(T, F, T, F)
c(T, T, F, F) | c(T, F, T, F)
# Which are big bills with a poor tip rate?
id <- (bill > 40 & rate < .10)
tips[id,]
```

## Your Turn

1. Read up on the dataset (`?diamonds`)
2. Plot price by carat (use qplot - go back to the motivating example for help with the syntax)
3. Create a variable `ppc` for price/carat. Store this variable as a column in the diamonds data
4. Make a histogram of all ppc values that exceed $10000 per carat.
5. Explore any other interesting relationships you find

### Solutions
2. Plot price by carat (use qplot - go back to the motivating example for help with the syntax)

```{r}
qplot(carat, price, data = diamonds)
```


3. Create a variable `ppc` for price/carat. Store this variable as a column in the diamonds data

```{r}
diamonds$ppc <- diamonds$price / diamonds$carat
```


4. Make a histogram of all ppc values that exceed $10000 per carat.

```{r}
qplot(ppc, geom = "histogram", data = diamonds[diamonds$ppc > 10000,])
```

5. Explore any other interesting relationships you find



# Data Structures

## Data Frames

- Data Frames are the work horse of R objects
- Structured by rows and columns and can be indexed
- Each column is a specified variable type
- Columns names can be used to index a variable
- Advice for naming variable applies to editing columns names
- Can be specified by grouping vectors of equal length as columns

Data frames can be indexed in several ways:

- Elements indexed similar to a vector using `[` `]`
- `df[i,j]` will select the element in the $i^{th}$ row and $j^{th}$ column
- `df[ ,j]` will select the entire $j^{th}$ column and treat it as a vector
- `df[i ,]` will select the entire $i^{th}$ row and treat it as a vector
- Logical vectors can be used in place of i and j used to subset the row and columns

New variables can be added to a data frame: 
1. Create a new vector that is the same length as other columns
2. Append new column to the data frame using the `$` operator
3. The new data frame column will adopt the name of the vector

### Example 
The following demonstration will use Edgar Anderson's Iris Data:

```{r}
flower <- iris
```

Select Species column (5th column):

```{r}
head(flower[,5])
```

Select Species column with the `$` operator:

```{r}
head(flower$Species)
```

Logical indexing: 
```{r}
head(flower$Species == "setosa")

head(flower[flower$Species=="setosa", ])
```

### Creating a Data Frame

Create a data frame using `data.frame` function

```{r}
mydf <- data.frame(NUMS = 1:5, 
                   lets = letters[1:5],
                   vehicle = c("car", "boat", "car", "car", "boat"))
mydf
```

Column names can be accessed with the `names` function. They can also be changed by assigning values to `names(df)`, as shown below. 

```{r}
# Use the `names` function to set that first column to lowercase:
names(mydf)[1] <- "nums"
mydf
```

### Your Turn

1. Make a data frame with column 1: 1,2,3,4,5,6 and column 2: a,b,a,b,a,b
2. Select only rows with value "a" in column 2 using logical vector
3. `mtcars` is a built in data set like `iris`: Extract the 4th row of the `mtcars` data.

#### Solutions

1. Make a data frame with column 1: 1,2,3,4,5,6 and column 2: a,b,a,b,a,b

```{r}
mydf <- data.frame(col1 = 1:6, col2 = rep(c("a", "b"), times = 3))

mydf
```

2. Select only rows with value "a" in column 2 using logical vector

```{r}
mydf[mydf$col2 == "a",]
```

3. Extract the 4th row of the `mtcars` data.

```{r}
data(mtcars)

mtcars[4,]
```



## Lists

- Lists are a structured collection of R objects
- R objects in a list need not be the same type
- Create lists using the `list` function
- Lists indexed using double square brackets `[[ ]]` to select an object

### Example

Create a list containing a matrix and a vector:

```{r}
mylist <- list(matrix(letters[1:10], nrow = 2, ncol = 5),
               seq(0, 49, by = 7))
mylist
```

Use indexing to select the second list element:

```{r}
mylist[[2]]
```

### Your Turn

1. Create a list containing a vector and a 2x3 data frame
2. Use indexing to select the data frame from your list
3. Use further indexing to select the first row from the data frame in your list

#### Solutions
1.  Create a list containing a vector and a 2x3 data frame

```{r}
mylist <- list(vec = 1:6, df = data.frame(x = 1:2, y = 3:4, z = 5:6))
```

2. Use indexing to select the data frame from your list

```{r}
mylist[[2]]
```

3. Select the first row from the data frame in your list

```{r}
mylist[[2]][1,]
```


## Examining Objects
R objects can get very complicated very quickly (nesting lists inside of lists, for example). 
The following functions are useful to figure out the structure of objects in R. 

- `head(x)` - View top 6 rows of a data frame
- `tail(x)` - View bottom 6 rows of a data frame
- `summary(x)` - Summary statistics   
- `str(x)` - View structure of object  
- `dim(x)` - View dimensions of object 
- `length(x)` - Returns the length of a vector

### Example

Examine the first two values of an object by passing the `n` parameter to the `head` function:

```{r}
head(diamonds, n = 2) # first 2 rows of diamonds data frame
tail(diamonds, n = 2) # last 2 rows of diamonds data frame
```

What's the structure of the object?

```{r}
str(diamonds) # structure of diamonds data frame
str(mylist) # structure of mylist list
```

How does R summarize objects?
```{r}
summary(diamonds) # summarize each column in diamonds
summary(mylist) # summarize mylist - # values in each item in the list
```

What are the dimensions of the object?
```{r}
dim(diamonds) # dimensions of diamonds data frame
dim(mylist) # mylist doesn't have dimensions because it isn't a rectangular object

length(diamonds) # diamonds is a data frame with 10 columns (or really, a list with 10 vectors that are the same length)
length(mylist) # mylist has 2 objects
```

### Your Turn

1. View the top 8 rows of mtcars data
2. What type of object is the mtcars data set?
3. How many rows are in iris data set? (try finding this using dim or indexing + length)
4. Summarize the values in each column in iris data set


#### Solutions
1.  View the top 8 rows of mtcars data

```{r}
head(mtcars, n = 8)
```

2. What type of object is the mtcars data set?

```{r}
str(mtcars)
```
  
3. How many rows are in iris data set? (try finding this using dim or indexing + length)

```{r}
dim(iris)
```

4. Summarize the values in each column in iris data set

```{r}
summary(iris)
```


## Working with Output from a Function

- Can save output from a function as an object
- Object is generally a list of output objects
- Can use items from the output for further computing
- Examine object using functions like `str(x)`

### Demo: Saving Output

Conduct a t-test using iris data to see if petal lengths for setosa and versicolor are the same. The `t.test` function can only handle two groups, so subset out the virginica species.

```{r}
iris.subset <- iris[iris$Species != "virginica", ]
t.test(Petal.Length ~ Species, data = iris.subset)
```

Save the output of the t-test to an object:

```{r}
tout <- t.test(Petal.Length ~ Species, data = iris.subset)
```

Look at the structure of the t-test object:

```{r}
str(tout)
```

Since this is simply a list, use regular indexing to access the p-value. 

```{r}
tout$p.value
tout[[3]]
```

## Importing Data

It is generally necessary to import data in to R rather than just using built-in datasets.

- Tell R where the data is saved using `setwd()` (or an appropriate file path)
- Read in data using R functions such as:
    - `read.table()` for reading in .txt files
    - `read.csv()` for reading in .csv files
    - the `readr` package has "smarter" versions of these functions and may be more useful
- Assign the data to new R object when reading in the file

### Importing Data Demo

First, create a csv file. We can use a text editor, excel... Then we load it in:

```{r, eval=FALSE}
littledata <- read.csv("PretendData.csv")
```

### Your Turn

- Make 5 rows of data in an excel spreadsheet and save it as a *tab-delimited txt file*.
- Import this new .txt file into R with `read.table`. You may need to look at the help page for `read.table` in order to properly do this.


#### Solutions
[Excel Spreadsheet](./data/Fun Webcomics.xlsx) (also at https://bit.ly/2hOYYYT)

```{r}
webcomics <- read.table("./data/FunWebcomics.txt", header = T)
webcomics
```


# Packages and Basic Programming


## R Packages

- Commonly used R functions are installed with base R
- R packages containing more specialized R functions can be installed freely from CRAN servers using function `install.packages()`
- After packages are installed, their functions can be loaded into the current R session using the function `library()`


### Finding R Packages

- How do I locate a package with the desired function?
- Google ("R project" + search term works well)
- R website task views to search relevent subjects: http://cran.r-project.org/web/views/
- ??searchterm will search R help for pages related to the search term
- `sos` package adds helpful features for searching for packages related to a particular topic

Handy R Packages:

- `ggplot2`: Statistical graphics
- `dplyr`/`tidyr`: Manipulating data structures
- `knitr`: integrate LaTeX, HTML, or Markdown with R for easy reproducible research


## Creating Functions

Code Skeleton:

```{r}
foo <- function(arg1, arg2, ...) {
    # Code goes here
    return(output)
}
```

Example:

```{r}
mymean <- function(data) {
    ans <- sum(data) / length(data)
    return(ans)
}
```


## If/Else Statements

Skeleton:

```{r, eval=FALSE}
if (condition) {
    # Some code that runs if condition is TRUE
} else {
    # Some code that runs if condition is FALSE
}
```

Example:

```{r}
mymean <- function(data) {
    if (!is.numeric(data)) {
        stop("Numeric input is required")
    } else {
        ans <- sum(data) / length(data)
        return(ans)
    }
}
```


## Looping

- Reduce the amount of typing
- Let R do repetitive tasks automatically
- R offers several loops: for, while, repeat.

```{r}
for (i in 1:3) {
    print(i)
}
```


### For Loops

```{r}
tips <- read.csv("https://bit.ly/2iNqvKM")

id <- c("total_bill", "tip", "size")
for (colname in id) {
    print(colname)
}

for(colname in id) {
    print(paste(colname, mymean(tips[, colname])))
}
```

### While Loops

```{r}
i <- 1
while (i <= 5) {
    print(i)
    i <- i + 1
}
```

## Your Turn 

1. Create a function that takes numeric input and provides the mean and standard deviation for the data (`sd` may be useful)
2. Add checks to your function to make sure the data is either numeric or logical.  If it is logical convert it to numeric.
3. Loop over the columns of the diamonds data set and apply your function to all of the numeric columns.

### Solutions 

1. Create a function that takes numeric input and provides the mean and standard deviation for the data (`sd` may be useful)
```{r}
myfun <- function(x) {
  m <- mean(x)
  s <- sd(x)
  return(c(mean = m, sd = s))
}
```

2. Add checks to your function to make sure the data is either numeric or logical.  If it is logical convert it to numeric.
```{r}
myfun <- function(x) {
  if (is.logical(x)) {
    x <- as.numeric(x)
  }
  if (!is.numeric(x)) {
    warning("x is not logical or numeric. Cannot compute a mean or std. deviation.")
    return(c(mean = NA, sd = NA))
  }
  m <- mean(x)
  s <- sd(x)
  return(c(mean = m, sd = s))
}
```

3. Loop over the columns of the diamonds data set and apply your function to all of the numeric columns.
```{r, warning = F}
data(diamonds)
diamondStats <- matrix(0, nrow = ncol(diamonds), ncol = 2, 
                       dimnames = list(names(diamonds), 
                                       c("mean", "sd")))

for(i in 1:ncol(diamonds)) {
  diamondStats[i,] <- myfun(diamonds[[i]])
}

diamondStats
```

