---
title: "Introduction to R"
output: 
    revealjs::revealjs_presentation:
        transition: fade
        theme: white
        highlight: zenburn
        css: ../css/slides.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T)
```

# Setting up R for Data Analysis

## Your Background

What do you know already?

> - Excel
> - SAS, SPSS, Minitab
> - Any other programming language
> - SQL or other database

## Installing R

For Windows or OS X:

- Go to http://www.r-project.org/
- Click the CRAN link on the left, and pick a download site (0-Cloud is a good choice)
- Choose link based on your OS
- On Windows, choose the "base" subdirectory to install R.
- On OS X, choose the .pkg file to install R.

## Installing RStudio

- Browse to https://www.rstudio.com/
- Mouse over Products and click RStudio
- Choose RStudio Desktop
- Click Download RStudio Desktop
- Choose the installer appropriate for your platform

# Example: Tips Data

## Goals

- Explore a real dataset using R
- Get the "flavor" of R for data management and exploration
- Don't focus on the code - it will be explained later and in much more detail

## Tips Dataset

A server recorded the tips they received over about 10 weeks, including several variables:

  - Amount they were tipped
  - Cost of the total bill
  - Characteristics about the party (# people, gender, etc.)

## Primary Question:

How do these variable influence the amount being tipped?

Follow along using [Tips-Example.R](code/Tips-Example.R)

## First Look at Data in R

Load the tips data using `read.csv()`

```{r, echo = T}
tips <- read.csv("http://srvanderplas.github.com/NPPD-Analytics-Workshop/01.Introduction/data/tips.csv")
# Short link:
tips <- read.csv("http://bit.ly/2fQoMP1")
```

The `head()` function shows the first few rows of the data: 
```{r, echo = T}
head(tips)
```

## Data Set Attributes

How big is the dataset? What types of variables are in each column?
```{r}
str(tips)
```

## Summary of Variables

R can easily summarize each variable in the dataset: 
```{r}
summary(tips)
```

## Plotting the data

First, we need to install and load `ggplot2`, a library for plotting the data
```{r, eval=FALSE}
install.packages("ggplot2")
library(ggplot2)
```

```{r, echo=FALSE}
library(ggplot2)
```

## Scatterplots

What is the relationship between total bill and tip value? 

```{r, fig.height=3, fig.width=7}
qplot(tip, total_bill, geom = "point", data = tips)
```

## Fancy Scatterplots

Color the points by meal. Is there a difference?

```{r, fig.height=4, fig.width=7}
qplot(tip, total_bill, geom = "point", data = tips, colour = time)
```

## Even More Scatterplots

Add a linear regression line to the plot

```{r, fig.height=4, fig.width=7}
qplot(tip, total_bill, geom = "point", data = tips) + 
    geom_smooth(method = "lm")
```


## Rate of Tipping

Tips are usually based on a percentage of the total bill. 

Make a new variable for the tipping rate = tip / total bill

```{r}
# New variable rate is a combination of 
# other variables in the tips dataset
tips$rate <- tips$tip / tips$total_bill

summary(tips$rate)
```

## Tipping Rate Histogram

What is the distribution of tipping rates? 

```{r, fig.height=4, fig.width=7}
qplot(rate, data = tips, binwidth = .01)
```

## Someone is an AMAZING tipper...

One person tipped over 70%, who are they?

```{r}
tips[which.max(tips$rate),]
```


## Rates by Gender

Look at the average tipping rate for men and women seperately

```{r}
mean(tips$rate[tips$sex == "Male"])
mean(tips$rate[tips$sex == "Female"])
```

## Statistical Significance

There is a difference but is it statistically significant?

```{r}
t.test(rate ~ sex, data = tips)
```

## Boxplots

Boxplots are useful for comparing the distribution of data. Do smokers tip at different rates than non-smokers?

```{r, fig.height=3, fig.width=7}
qplot(smoker, rate, geom = "boxplot", data = tips)
```

## Your Turn {data-background=#527a7a}

Try playing with chunks of code from this session to further investigate the tips data:

1. Get a summary of the total bill values
2. Make side by side boxplots of tip rates for different days of the week
3. Find the average tip value for smokers

## Your Turn Solutions
### Summary  of Total Bill Values
```{r}
summary(tips$total_bill)
```

### Tip Values by Day of the Week

```{r, fig.height=3, fig.width=7}
qplot(day, rate, geom = "boxplot", data = tips)
```

### Average Tip Value for Smokers
```{r}
mean(tips$tip[tips$smoker == "Yes"])
```
